MonteCarlo Simulation's of the Covid-19 Pandemic

### 2022-12-15 paper.rmd
Added China portion of paper

---
title: "MonteCarlo Simulation's of the Covid-19 Pandemic"
author: "Austin Lackey, Chien Lin and Jin Peng"
date: "Tuesday, December 6th, 2022"
output: 
  pdf_document:
    toc: TRUE
    toc_depth: 2
    number_sections: TRUE
    df_print: paged
    fig_caption: TRUE
    highlight: tango
---
```{r setup, include=FALSE}
# useful libraries
library(tidyverse)
library(pracma)
library(minpack.lm)
# read covid data
data <- read_csv("owid-covid-data.csv")
set.seed(420)
knitr::opts_chunk$set(fig.width=3, fig.height=3) 
```

@@

# recreate plots
positive <- ggplot(data = china) +
  geom_point(aes(x = date, y = total_cases)) +
  theme(title =element_text(size=6)) +
  labs(x = "Date", y = "China cumulative number of cases")
positive
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold", out.width="50%"}
ggplot(data = china) +
  geom_point(aes(x = date, y = total_deaths)) +
  theme(title =element_text(size=6)) +
  labs(x = "Date", y = "China cumulative number of fatalities") +
  ylim(0,5000)
```

@@

In the study, to fit the data, the authors stated that they used a function of the type of the Gauss error function, that is, the integral of a normal, Gaussian distribution: 
$$a+b*erf(c*x-d)$$ 
There are four parameters $a,b,c,d$ in this model, which shows great level of fit. Since they didn't specify how to realize the function in R, we went around and found that the $nls$ function did a great job on what we need. We took a cheap gauss of $a,b,c,d$ from the preview of the data as the start points, and apply the $nls$ function to determine the nonlinear (weighted) least-squares estimates of the four parameters $a,b,c,d$. Here is an example code of the model construction by the $nls$ function:

@@

```{r, results=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
gauss_fn <- function(x,a,b,c,d) {a+b*erf(c*x-d)}
model1 <- nls(total_cases ~ gauss_fn(days,a,b,c,d), data=china, start=list(a=10000,b=10000,c=0.1,d=1))
# summary(model1)$parameters
fn_model1 <- function(x) {4.017e+04 + 4.047e+04 * erf(9.284e-02 * x - 1.732e+00)}
ggplot(data = china) +
  geom_point(aes(x = days, y = total_cases)) +
  theme(title =element_text(size=6)) +
  labs(x = "Days", y = "China cumulative number of cases") +
  stat_function(fun=fn_model1, color="#d33636", size=1.5) +
  geom_vline(xintercept = peak1, color="#d33636", size=1.5, linetype="dashed")
```

```{r, results=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
try <- china$total_deaths
model_try <- nls(try ~ gauss_fn(days,a,b,c,d), data=china, start=list(a=2200,b=2200,c=0.1,d=1))
est_try <- summary(model_try)$parameters
est_try <- est_try[, 1]
names(est_try) <- NULL
fn_try <- function(x, ests) {ests[1]  + ests[2] * erf(ests[3] * x - ests[4])}

# find the second derivative of fitted function
# source: https://proofwiki.org/wiki/Derivative_of_Error_Function
erf_deriv <- function(x) {2/sqrt(pi)*exp(-x^2)}
fn_try_deriv <- bquote(.(est_try[2]) * 2/sqrt(pi) * exp(-(.(est_try[3]) * x - .(est_try[4]))^2))
fn_try_deriv <- D(fn_try_deriv,'x')
x=seq(1,67,length.out=6700)
tmp <- eval(fn_try_deriv)
peak2 <- min(x[tmp < 0])
```

```{r, results=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
model2 <- nls(total_deaths ~ gauss_fn(days,a,b,c,d), data=china, start=list(a=10000,b=10000,c=0.1,d=1))
#summary(model2)$parameters
fn_model2 <- function(x) {1.571e+03 + 1.658e+03 * erf(6.194e-02 * x - 1.533e+00)}
ggplot(data = china) +
  geom_point(aes(x = days, y = total_deaths)) +
  theme(title =element_text(size=6)) +
  labs(x = "Days", y = "China cumulative number of fatalities") +
  ylim(0,5000) +
  stat_function(fun=fn_model2, color="#1b95ff", size=1.5) +
  geom_vline(xintercept = peak2, color="#1b95ff", size=1.5, linetype="dashed")
```

@@

>>>>

---
title: "MonteCarlo Simulation's of the Covid-19 Pandemic"
author: "Austin Lackey, Chien Lin and Jin Peng"
date: "Tuesday, December 6th, 2022"
output:
  pdf_document:
    toc: yes
    toc_depth: 2
    number_sections: yes
    df_print: paged
    fig_caption: yes
    highlight: tango
  html_document:
    toc: yes
    toc_depth: '2'
    df_print: paged
---
```{r setup, include=FALSE}
# useful libraries
library(tidyverse)
library(pracma)
library(minpack.lm)
# read covid data
data <- read_csv("owid-covid-data.csv")
set.seed(420)
knitr::opts_chunk$set(fig.width=3, fig.height=3) 
```

@@

# recreate plots
positive <- ggplot(data = china) +
  geom_point(aes(x = date, y = total_cases)) +
  theme(title =element_text(size=6)) +
  labs(x = "Date", y = "China cumulative number of cases")
positive

ggplot(data = china) +
  geom_point(aes(x = date, y = total_deaths)) +
  theme(title =element_text(size=6)) +
  labs(x = "Date", y = "China cumulative number of fatalities") +
  ylim(0,5000)
```

@@

In the study, to fit the data, the authors stated that they used a function of the type of the Gauss error function, that is, the integral of a normal, Gaussian distribution: 
$$a+b*erf(c*x-d)$$

There are four parameters $a,b,c,d$ in this model, which shows great level of fit. Since they didn't specify how to realize the function in R, we went around and found that the $nls$ function did a great job on what we need. 
We took a cheap gauss of $a,b,c,d$ from the preview of the data as the start points, and apply the $nls$ function to determine the nonlinear (weighted) least-squares estimates of the four parameters $a,b,c,d$. 
Here is an example code of the model construction by the $nls$ function:

@@

```{r, results=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
gauss_fn <- function(x,a,b,c,d) {a+b*erf(c*x-d)}
model1 <- nls(total_cases ~ gauss_fn(days,a,b,c,d), data=china, start=list(a=10000,b=10000,c=0.1,d=1))
# summary(model1)$parameters
fn_model1 <- function(x) {4.017e+04 + 4.047e+04 * erf(9.284e-02 * x - 1.732e+00)}
ggplot(data = china) +
  geom_point(aes(x = days, y = total_cases)) +
  theme(title =element_text(size=6)) +
  labs(x = "Days", y = "China cumulative number of cases") +
  stat_function(fun=fn_model1, color="#d33636", size=1.5)

try <- china$total_deaths
model_try <- nls(try ~ gauss_fn(days,a,b,c,d), data=china, start=list(a=2200,b=2200,c=0.1,d=1))
est_try <- summary(model_try)$parameters
est_try <- est_try[, 1]
names(est_try) <- NULL
fn_try <- function(x, ests) {ests[1]  + ests[2] * erf(ests[3] * x - ests[4])}

# find the second derivative of fitted function
# source: https://proofwiki.org/wiki/Derivative_of_Error_Function
erf_deriv <- function(x) {2/sqrt(pi)*exp(-x^2)}
fn_try_deriv <- bquote(.(est_try[2]) * 2/sqrt(pi) * exp(-(.(est_try[3]) * x - .(est_try[4]))^2))
fn_try_deriv <- D(fn_try_deriv,'x')
x=seq(1,67,length.out=6700)
tmp <- eval(fn_try_deriv)
peak2 <- min(x[tmp < 0])

model2 <- nls(total_deaths ~ gauss_fn(days,a,b,c,d), data=china, start=list(a=10000,b=10000,c=0.1,d=1))
#summary(model2)$parameters
fn_model2 <- function(x) {1.571e+03 + 1.658e+03 * erf(6.194e-02 * x - 1.533e+00)}
ggplot(data = china) +
  geom_point(aes(x = days, y = total_deaths)) +
  theme(title =element_text(size=6)) +
  labs(x = "Days", y = "China cumulative number of fatalities") +
  ylim(0,5000) +
  stat_function(fun=fn_model2, color="#1b95ff", size=1.5)
```

@@

The day of flex is the peak day of daily cases, that is, the day that the second derivative of the fit of cumulative number changes sign, from acceleration to deceleration. 
To analyze the day of flex of positive cases and fatalities in China, we took the second derivative of the Gauss error function we fitted above, and used the first day with negative second derivative as the day of flex. 
The red dashed vertical line shows the day of flex.

```{r, results=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
gauss_fn <- function(x,a,b,c,d) {a+b*erf(c*x-d)}
model1 <- nls(total_cases ~ gauss_fn(days,a,b,c,d), data=china, start=list(a=10000,b=10000,c=0.1,d=1))
# summary(model1)$parameters
fn_model1 <- function(x) {4.017e+04 + 4.047e+04 * erf(9.284e-02 * x - 1.732e+00)}
ggplot(data = china) +
  geom_point(aes(x = days, y = total_cases)) +
  theme(title =element_text(size=6)) +
  labs(x = "Days", y = "China cumulative number of cases") +
  stat_function(fun=fn_model1, color="#d33636", size=1.5)

model2 <- nls(total_deaths ~ gauss_fn(days,a,b,c,d), data=china, start=list(a=10000,b=10000,c=0.1,d=1))
#summary(model2)$parameters
fn_model2 <- function(x) {1.571e+03 + 1.658e+03 * erf(6.194e-02 * x - 1.533e+00)}
ggplot(data = china) +
  geom_point(aes(x = days, y = total_deaths)) +
  theme(title =element_text(size=6)) +
  labs(x = "Days", y = "China cumulative number of fatalities") +
  ylim(0,5000) +
  stat_function(fun=fn_model2, color="#1b95ff", size=1.5) +
  geom_vline(xintercept = peak2, color="#1b95ff", size=1.5, linetype="dashed")
```

The day of flex of cumulative number of diagnosed positive cases in China is 18.66 and the day of flex of cumulative number of fatalities in China is 24.74. 
We can tell that they are both around the middle of the climbing portion of the data, where the cumulative number starts to decelerate in growing.
